---
title: "Body measurements"
output: html_document
---

## imports
```{r}
#imports
```

##global variables

```{r}
quantile_cutoff <- 0.9

```

## Data reading

Data was downloaded from ..... and contains these columns:

## Format
A data frame with 252 observations on the following 19 variables.

- `case` - Case Number
- `body.fat` - Percent body fat using Brozek's equation: 457/Density - 414.2
- `body.fat.siri` - Percent body fat using Siri's equation: 495/Density - 450
- `density` - Density (gm/cm\^2)
- `age` - Age (yrs)
- `weight` - Weight (lbs)
- `height` - Height (inches)
- `BMI` - Adiposity index = Weight/Height^2 (kg/m^2)
- `ffweight` - Fat Free Weight = (1 - fraction of body fat) * Weight, using Brozek's formula (lbs)
- `neck` - Neck circumference (cm)
- `chest` - Chest circumference (cm)
- `abdomen` - Abdomen circumference (cm) "at the umbilicus and level with the iliac crest"
- `hip` - Hip circumference (cm)
- `thigh` - Thigh circumference (cm)
- `knee` - Knee circumference (cm)
- `ankle` - Ankle circumference (cm)
- `bicep` - Extended biceps circumference (cm)
- `forearm` - Forearm circumference (cm)
- `wrist` - Wrist circumference (cm) "distal to the styloid processes"


```{r}

body_measurements <- read.table(file = "body_measurements.csv",
           header = TRUE,
           sep = ";",
           dec = ",",
           na.strings = "-99")

str(body_measurements)
```

