---
title: "DAVuR2"
date: '2022-04-25'

output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "> ")
#knitr::opts_chunk$set(warning = FALSE)
library(ggplot2)
```

# Look at the data

```{r, echo = FALSE}
knitr::kable(head(airquality))
```

## first figure

```{r airquality-plot, warning=FALSE, message=FALSE, fig.cap="Ozone depending on solar radiotoion for the different months"}
airquality$Month_f <- factor(airquality$Month, levels = 5:9, labels = month.abb[5:9])

ggplot(data = airquality, 
       mapping = aes(x = Solar.R, y = Ozone)) +
    geom_smooth(mapping = aes(color = Month_f, linetype = Month_f), method = "lm", se = FALSE, size = 0.5) +
    geom_point(size = 0.8, alpha = 0.5, shape = 9) +
    xlab("Solar Radiation (lang)") +
    ylab("Ozone (ppb)") +
    facet_wrap(. ~ Month_f)
```

# Chickens

Select

```{r, comment="*"}
#subset(x = ChickWeight, subset = Chick %in% 1:3)
#ChickWeight[ChickWeight$Chick %in% 1:3, ]

library(dplyr)
first_three_chicks <- filter(ChickWeight, Chick %in% 1:3)
```

Plot:

```{r}
ggplot(data = first_three_chicks,
       mapping = aes())
```

```{r}
#install.packages("ggpubr")
library(ggpubr)
ggplot(data = na.omit(airquality), 
       mapping = aes(x = Wind, y = Ozone, color=Solar.R)) +
    geom_point(size = 2, alpha = 0.8) +
    scale_color_gradient(low = "#fff021", high = "#a30300") +
    theme_pubr(legend = "right")
```

```{r}
sp <- ggplot(faithful, aes(x=eruptions, y=waiting)) 
sp + 
    geom_point(alpha = 0.6) + 
    geom_density_2d()
```

```{r}
ggplot(faithful, aes(y=waiting, x = "")) +
    geom_violin()

```

```{r}
library(ggplot2)
library(dplyr)

x <- rnorm(100)
hist(x, breaks = 100)
df <- data.frame(x = x)
ggplot(data = df, mapping = aes(sample = x)) +
    geom_qq() + stat_qq_line()

shapiro.test(x)

airquality %>% ggplot(mapping = aes(sample = Solar.R)) +
    geom_qq() + stat_qq_line()

airquality %>% ggplot(mapping = aes(x = Solar.R)) +
    geom_histogram()

shapiro.test(airquality$Solar.R)
```

```{r}
airquality$Month <- factor(airquality$Month)
ggplot(data = airquality, mapping = aes(x = Solar.R, y = Ozone, linetype = Month)) +
    geom_smooth(method="lm") +
    geom_point(aes(color = Month), alpha = 0.7) +
    facet_wrap(. ~ Month, ncol = 1)
```

```{r}
ggplot(data = cars, mapping = aes(x = speed, y = dist)) + 
    geom_point() +
    xlim(0, 30) +
    scale_y_log10() + 
    scale_x_reverse() +
    theme_minimal(base_size = 14)
```

## Exploratie Broodbeleg

```{r}
url <- "https://raw.githubusercontent.com/MichielNoback/datasets/master/Broodbeleg/Broodbeleg_2.txt"
broodbeleg <- read.table(file = url, sep = "\t", header = T)
#str(broodbeleg)
#head(broodbeleg)
```

```{r}
protein_f <- cut(x = broodbeleg$eiwit, 
    breaks = c(0, 2, 20, 100), 
    labels = c("low prot", "medium prot", "high prot"),
    right = F)
table(protein_f)
broodbeleg$protein_f <- protein_f
#head(broodbeleg)

library(ggrepel)
ggplot(data = broodbeleg,
       mapping = aes(x = vet.tot, 
                     y = H.kcal, 
                     color = kool.tot)) +
    geom_point(mapping = aes(size = eiwit), alpha = 0.6) +
    geom_text_repel(mapping = aes(label = Omschrijving), hjust=0, vjust=0, size = 2) + 
    scale_color_gradient(low = "blue", high = "darkred") +
    facet_wrap(. ~ protein_f)

```

## Tidyr

```{r}
library(tidyr)
set.seed(1234)
two_two_col <- tibble(subject = letters[1:5],
              T0_Control = rnorm(5, 10, 1),
              T0_Treated = rnorm(5, 9.5, 1.2),
              T1_Control = rnorm(5, 11, 1.5),
              T1_Treated = rnorm(5, 16, 2))
two_two_col
```

```{r}
pivot_longer(data = two_two_col,
             cols = 2:5,
             names_pattern = "(T\\d)_(Treated|Control)",
             names_to = c("Time", "Treated"),
             values_to = "Response")
```

```{r}
untidy2 <- read.table(
  file = "https://raw.githubusercontent.com/MichielNoback/davur1_2022/master/untidy2.csv",
  header = T, 
  sep = ",")

(untidy2 <- as_tibble(untidy2))

library(tidyr)
tidy2 <- pivot_longer(data = untidy2,
             cols = 2:3,
             names_to = "Time",
             values_to = "Creatine",
             names_pattern = "creatine_(T.)")
```

## plot with strip chart

```{r}
ggplot(data = tidy2,
       mapping = aes(x = Time, y = Creatine)) +
    geom_point() +
    geom_text(mapping = aes(label = patient), hjust = -1)
```

## untidy1

```{r}
untidy1 <- read.table(file = "https://raw.githubusercontent.com/MichielNoback/davur1_2022/master/untidy1.csv",
           head = T,
           sep = ",")
(untidy1 <- as_tibble(untidy1))
pivot_wider(data = untidy1, 
            names_from = type,
            values_from = value)
```

```{r}
# library
library(ggplot2)
library(ggExtra)
 
# The mtcars dataset is proposed in R
head(mtcars)
 
# classic plot :
p <- ggplot(data = mtcars, 
            aes(x=wt, y=mpg, color=cyl, size=cyl)) +
      geom_point() +
      theme(legend.position="none")
 
# marginal density
ggMarginal(p, type="density")
```

```{r}
airquality %>% 
    filter(Month %in% 6:7) %>% 
    select(Wind, Temp, Month, Day) %>% 
    group_by(Month) %>%
    summarize(mw = mean(Wind), 
              mt = mean(Temp))


aggregate(cbind(Wind, Temp) ~ Month, 
          data = airquality[airquality$Month %in% 6:7, 3:6], 
          FUN = mean)


aggregate(cbind(Wind, Temp) ~ Month, 
          data = subset(x = airquality, 
                        subset = Month %in% 6:7, 
                        select = 3:6), 
          FUN = mean)
```

```{r}
ggplot(data = Orange, 
       mapping = aes(x = age, y = circumference, linetype = Tree)) +
  geom_line()
```

```{r}
library(dplyr)
# iris %>%
#   group_by(Species) %>%
#   ##same: summarize(across(everything(), mean)) %>%
#   ##same: summarize(across(1:4, mean)) %>%
#   summarize(across(where(is.numeric), mean)) %>%
#   pivot_longer(cols = -Species, names_to = "Part", values_to = "Mean")

means <- mtcars %>%
  group_by(cyl, carb) %>%
  summarize(mpg = mean(mpg), .groups = "drop") %>%
  mutate(cyl = as.factor(cyl), carb = as.factor(carb)) %>%
  complete(cyl, carb)
means

ggplot(means, aes(x = cyl, y = carb, fill = mpg)) + 
  geom_tile(width = .95, height = 0.95) + 
  scale_fill_viridis_c(option = "B", begin = 0.15, end = 0.98,
                       name = "mpg") + 
  scale_y_discrete() +
  coord_fixed(expand = FALSE) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_blank(),
        #axis.text.y = element_text(size = 14),
        legend.title = element_text(size = 12))
```

## scales

```{r}
cars %>% ggplot(mapping = aes(x = speed, y = dist)) +
  geom_point() +
  coord_polar()


airquality %>% 
  mutate(date = lubridate::ymd(paste("1973", Month, Day, sep = "-"))) %>%
  ggplot(aes(x = date, y = Temp)) + 
  geom_point() +
  geom_smooth(se = F) +
  coord_polar() +
  scale_y_continuous(trans = "log2") +
  theme_minimal()
```

## Titanic

```{r}
titanic_tbl <- as.data.frame(Titanic) 
titanic_tbl <- titanic_tbl %>%
    filter(Class != "Crew") %>%
    as_tibble(titanic_tbl) %>%
    group_by(Class, Sex) %>%
    arrange(Class, desc(Sex)) %>%
    summarise(counts = sum(Freq), .groups = "drop") %>% 
    mutate(Class = paste(Class, "class"),
           Sex = factor(Sex, levels = c("Female", "Male"))) %>%
    group_by(Class) %>%
    mutate(nlabel = cumsum(counts) - counts/2)
    
titanic_tbl

titanic_tbl %>% ggplot(mapping = aes(x = Class, y = counts, fill = Sex)) +
    geom_col(position = "stack", color = "white", size = 1, width = 1) +
    geom_text(aes(y = nlabel, label = counts), color = "white", size = 4) +
    scale_x_discrete(expand = c(0, 0), name = NULL) +
    scale_y_continuous(expand = c(0, 0), breaks = NULL, name = NULL) +
    scale_fill_manual(
        values = c("#D55E00", "#0072B2"),
        breaks = c("Female", "Male"),
        labels = c("female passengers   ", "male passengers"),
        name = NULL) +
    theme(
        panel.grid.major = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(size = 12),
        legend.position = "bottom",
        legend.justification = "center",
        legend.background = element_rect(fill = "white"),
        legend.spacing.x = grid::unit(4.5, "pt"),
        legend.spacing.y = grid::unit(0, "cm"),
        legend.box.spacing = grid::unit(7, "pt"),
        panel.background = element_blank()
    )

```

# dplyr

Use case: selecting rows where *any* of the variables contains NA

```{r}
airquality %>% 
    filter(if_any(where(is.numeric), is.na))

airquality %>% 
    filter(if_any(everything(), is.na))

airquality %>% 
    filter(if_any(2:4, is.na))
```

# extract data from plot

```{r}
p <- ggplot(data=mtcars, mapping = aes(x = hp, y = wt)) + 
    geom_point() +
    geom_smooth(formula = y ~ x, method = "lm")
str(ggplot_build(p)$data)

p
```

# across

```{r}
iris %>% mutate(across(.cols = 1:2, 
                       .fns = ~ log2(.)))

iris %>% mutate(across(.cols = matches("\\."), 
                       .fns = ~ log2(.)))

iris %>% mutate(across(.cols = starts_with("P"), 
                       .fns = ~ log2(.)))

iris %>% mutate(across(.cols = where(is.numeric), 
                       .fns =  list(~ log2(.), ~ exp(.))))
```

```{r}
airquality <- airquality %>%
    mutate(temp_f = cut(x = Temp, 
                        breaks = c(50, 65, 85, 100), 
                        labels = c("low", "mid", "high")))

airq_mut <- airquality %>% 
    mutate(
        month_f = factor(x = Month, 
                     levels = 4:9, 
                     labels = c("april", "mei", "juni", "juli", "aug", "sept")),
        row_count = 1:nrow(airquality))

ggplot(data = airq_mut,
       mapping = aes(x = Ozone, y = Solar.R, size = Temp)) +
    geom_point(mapping = aes(color = month_f)) +
    geom_smooth(se = FALSE, color = "darkgreen")


ggplot(data = airq_mut,
       mapping = aes(x = month_f, y = Solar.R)) +
    geom_violin() #+
   # geom_boxplot()

```

## Select

```{r}
airquality %>%
    arrange(desc(Solar.R)) %>%
    slice(1:3)
#    top_n(3, Solar.R)

```


## sample_XX


```{r}
set.seed(1234)
sample_frac(tbl = airquality, size = 0.1)

```


```{r}
weights <- ChickWeight$weight
weight_change = numeric(length(weights))
for(i in 1:length(weights)) {
    weight_change[i] <- weights[i+1] - weights[i]
}
weight_change
```



```{r}
weights <- ChickWeight$weight
weights - lag(weights)
```

```{r}
(airq <- as_tibble(airquality))

(airq <- airq %>%
    mutate(Temp_change = Temp - lag(Temp)))
```
# Distributies 

- histogram
- density
- boxplot
- violin

```{r}
ggplot(data = airq, mapping = aes(x = Temp_change, color = factor(Month))) +
    geom_histogram() +
    facet_wrap(. ~ Month)
```


```{r}
#~ mean(., trim = 0.2
#function(x) mean(x, trim = 0.2)

f <- function(x) mean(x, trim = 0.2)
f(rnorm(50))


```



```{r}
my_z <- function(x) {
    abs((abs(x - mean(x)) / sd(x)))
}

iris %>% 
    mutate_at(vars(contains(".")), my_z) %>% 
    head(2)

```



```{r}
as_tibble(airquality) %>%
    mutate(Month = recode(Month, `5` = "Mei"))

iris %>% mutate(sp = recode(Species, viriginica = "Vir"))

recode(airquality$Month, `5` = "Mei")
```


# Conditioneel kolom select

Selecteer daar waar ten minste een kolom aan een voorwaarde voldoet

```{r}
rowAny <- function(x) rowSums(x) > 0 
airquality %>% 
    filter(
        rowAny(
            across(
                .cols = everything(),
                .fns = ~ is.na(.x)
            )
        )
    )


airquality %>% 
    filter(if_any(where(is.numeric), is.na))

airquality %>% 
    filter(if_any(everything(), is.na))

airquality %>% 
    filter(if_any(2:4, is.na))

```




```{r}
iris %>% 
    mutate_at(vars(contains(".")), my_z) %>% 
    head(2)

iris %>% mutate(across(contains("."), my_z))

iris %>% mutate(sepal_l_z = my_z(Sepal.Length),
                sepal_w_z = my_z(Sepal.Width))

```

## Groups

base R

```{r}
sapply(X = split(iris, f = iris$Species), FUN = function(x) mean(x$Sepal.Width))
```


dplyr

```{r}
iris %>% 
    group_by(Species) %>%
    summarize(m = mean(Sepal.Width))
```

```{r, message=FALSE}
ChickWeight %>%
    group_by(Diet, Time) %>%
    summarize(mean_weight = mean(weight), .groups = "drop") %>%
    ungroup() %>%
    ggplot(mapping = aes(x = Time, y = mean_weight, color = Diet)) +
        geom_point() +
        geom_line() +
        labs(x = "Time (days)",
             y = "Mean weigth (grams)") +
    theme_minimal()

```


```{r}
ChickWeight %>% 
    group_by(Chick) %>% #split on chicks
    mutate(rel_weight_gain = (weight - lag(weight))/weight) %>% 
    ungroup() %>% ## not required
    group_by(Diet) %>%
    summarize(mean_rel_gain = mean(rel_weight_gain, na.rm = T), 
              sd = sd(rel_weight_gain, na.rm = T))
```


```{r}
trA <- tribble(
  ~ID, ~var1,
  "a",   4,
  "b",   6,
  "c",   1,
)

trB <- tribble(
  ~id, ~var2,
  "a",   7,
  "b",   3,
  "d",   5
)


full_join(trA, trB, by = c("ID" = "id"))
inner_join(trA, trB, by = c("ID" = "id"))
left_join(trA, trB, by = c("ID" = "id"))
right_join(trA, trB, by = c("ID" = "id"))

trC <- tribble(
  ~ID, ~var1,
  "g",   7,
  "h",   5,
  "i",   6
)

bind_rows(trA, trC)
bind_cols(trA, trB)

```



```{r}
url <- "https://raw.githubusercontent.com/MichielNoback/datasets/master/rubella_measles_cases/measles_cases_by_country_by_month.csv"

measles <- read.table(file = url,
                      header = T,
                      sep = ";")


library(dplyr)
library(tidyr)
measles_long <- measles %>%
    select(-X) %>%
    pivot_longer(cols = 5:16, 
                 names_to = "month",
                 values_to = "count") %>%
    mutate(month = factor(month, levels = month.name))

head(measles_long)
(tmp <- measles_long %>% 
    group_by(Region, month) %>%
    summarise(month_totals = sum(count, na.rm = T), .groups = "drop"))

tmp %>%
    ggplot(mapping = aes(x = month)) +
        geom_bar(aes(weight = month_totals, fill = Region), 
                 position = "dodge") +
        theme(axis.text.x = element_text(angle = 90))


```

- *SEAR*: South-East Asia Region -- Bangladesh, Bhutan, Democratic People's Republic of Korea, India, Indonesia, Maldives, Myanmar, Nepal, Sri Lanka, Thailand, Timor
- *AFR*: Africa
- *WPR*: Western Pacific Region -- a.o. Mongolia, China, Japan, Papua New Guinea, Malaysia, Philippines, Vietnam, South Korea


## Date / time data


```{r}
ts <- read.table("~/Desktop/time_series_drug_test.txt", 
                 header = T,
                 sep = ";",
                 as.is = 2:3)

ts <- ts %>% mutate(
    date_time = paste(date, time),
    dt = lubridate::dmy(date),
    dttm = lubridate::dmy_hm(date_time)
)

ggplot(data = ts, 
       mapping = aes(x = dttm, y = response, linetype = subject, color = control)) +
    geom_line()
```


## RE

```{r}
string <- "I love the series 'allo 'allo!!, which means Hallo, Hallo!"
stringr::str_view_all(string,  pattern = "([Hh']?allo,? ?){1,2}\\!+")
```

```{r}
library(stringr)
str_to_title(string)
```



```{r}
paste0("The value of pi is ", pi, " and the first month of the year is ", month.name[1])

str_glue("The value of pi is {pi} and the first month of the year is {month.name[1]}")

```


## Protein analysis

```{r}
hgbA <- ">tr|A0A222AH57|A0A222AH57_CAVAP Hemoglobine alpha globin subunit OS=Cavia aperea OX=37548 GN=HBA PE=3 SV=1
MVLSTADKNNVKTTWDKIGGHAPEYVAEGLTRMFTSFPTTKTYFHHIDVSPGSGDIKAHG
KKVADALTTAVGHLDDLPTALSTLSDVHAHKLRVDPVNFKFLNHCLLVTLAAHLGADFTP
SIHASLDKFFASVSTVLTSKYR"

#stringr::str_extract()

header <- str_split(hgbA, "\n")[[1]][1]
#of
header <- str_trim(str_extract(hgbA, "^.+\n"))

split_fasta <- str_split(hgbA, "\n")[[1]]
seq <- split_fasta[2:length(split_fasta)]

seq <- str_c(seq, collapse = "")
seq

str_view_all(header, "OS=[A-Z][:lower:]+ [:lower:]+")
str_extract(header, "OS=[A-Z][:lower:]+ [:lower:]+")
organism <- str_match(header, "OS=([A-Z][:lower:]+ [:lower:]+)")[, 2]
organism


str_view_all(header, ">[:alpha:]+\\|[:alnum:]+\\|[a-zA-Z0-9_]+")


```




# Belangrijkste functies om te beheersen

## base R
-   cut()
-   factor()
-   c()
-   data.frame()
-   split()
-   apply()
-   nrow(), ncol()
-   names()
-   seq()
-   

## dplyr::
- select
- filter
- mutate & mutate(across())
- arrange
- group_by      
- summarize (summarise)

