---
title: "R Notebook"
output: html_notebook
---


```{r}
body_measurements <- read.table(file = "body_measurements.csv",
           header = TRUE,
           sep = ";",
           dec = ",",
           na.strings = "-99")

str(body_measurements)
```


# Explore age variable

```{r}
body_measurements$age
max(body_measurements$age, na.rm = TRUE)
min(body_measurements$age)
range(body_measurements$age)
mean(body_measurements$age)
median(body_measurements$age)
sd(body_measurements$age)
summary(body_measurements$age)
```

```{r}
age <- body_measurements$age
x2 <- seq(from = min(age), to = max(age), length = 100)
fun <- dnorm(x2, mean = mean(age), sd=sd(age))
hist(age,
     breaks = 25,
     main = "Distribution of age",
     xlab = "Age (years)",
     prob = TRUE,
     col = "gold")
lines(x = x2, 
      y = fun,
      col = "darkblue", 
      lwd=2)
lines(density(age), 
      col = "darkred", 
      lwd = 2)
```
## QQ plot

```{r}
qqnorm(age, pch = 20, cex = 0.5, col = rgb(1, 0, 1, 0.5))
qqline(age, col = "steelblue", lwd = 2)
```

## test for normality


```{r}
shapiro.test(rnorm(1000))
shapiro.test(age)
```


## Accessing elements from vector

```{r, results = "hold"}
x <- c(2.4, 3.5, 6.2, 7.2, 2.9, 5.4, 7.2)

##BY INDEX
#single index
x[2]
#series of indices
x[2:4]
#multiple/random indices
x[c(1, 3, 3, 2)]

##BY LOGICAL
x[c(T, F, F, T, F, F, T)]
x[c(T, F)] #will cycle the logical - every second element 

## BY BOOLEAN EXPRESSION
#this is a boolean that will generate a T/F for every element of x
x < 6

#you can use this in many ways
x[x < 6]
x[x != 7.2]
x[x < 6 & x > 3]
x[x < 3 | x > 6]
x[x**2 < 25]
```

**Mastering this is _essential_ to mastering R**

Not only is it important when working with vectors, but also with dataframes. 
The difference is that with dataframes there is an extra dimension.


```{r}
#body_measurements
with(body_measurements,
     {
       plot(
         x = age,
         y = BMI,
         pch = 20,
         #plotting character/symbol
         #col = "darkblue", #color of plot symbol
         col = rgb(
           red = 0.1,
           green = 0,
           blue = 1,
           alpha = 0.7
         ),
         cex = 0.7,
         #size of plot symbol
         xlab = "age (years)",
         ylab = "BMI (kg/m^2)"
       )
       #add regression line
       my_lm <- lm(BMI ~ age)
       abline(my_lm)
       my_lm
     })


```

# Genetic disoirders

```{r}
getwd()
disorders <- read.csv(file = "genetic_disorders/test.csv",
         na.strings = c("", "-99"))
dim(disorders)
```
Plot age versus

```{r}
plot(x = disorders$Patient.Age, 
     y = disorders$Blood.cell.count..mcL.,
     pch = 20,
     cex = 0.5,
     col = rgb(0, 0, 1, alpha = 0.4))
```
Better with boxplot, using *formula notation*

```{r}

boxplot(disorders$Blood.cell.count..mcL. ~ disorders$Patient.Age,
        xlab = "Patient age (years)",
        ylab = "Blood cell count (M/cl)")
```


Without formula

Dit is heel erg slecht!

```{r}
boxplot(body_measurements$age, body_measurements$BMI)
```
Beter is gebruikmaken van `par()`


```{r}
par(mfrow = c(1, 2))
boxplot(body_measurements$age, ylab = "age (years)")
boxplot(body_measurements$BMI, ylab = "BMI (kg/m^2)")

par(mfrow = c(1, 1))
```


# Creating factors (grouping variable)

```{r}
age <- c(body_measurements$age)
age_f <- cut(x = age,
    breaks = c(20, 35, 60, 100), 
    labels = c("young_adult", "adult", "senior"),
    right = FALSE)
body_measurements$age_group <- age_f
```

Use to create summary 

```{r}
boxplot(body_measurements$abdomen ~ body_measurements$age_group )
```

## Other way to create factors

```{r}
eyes <- c("brown", "blue", "blue", "blue", "brown", "brown")
class(eyes)
eye_f <- factor(eyes, levels = c("blue", "brown", "green"))
barplot(table(eyes))

dog_size <- c("tiny", "large", "normal", "large", "large", "enormous", "normal", "tiny", "normal", "normal")
(dog_size_f <- factor(x = dog_size,
                      levels = c("tiny", "normal", "large", "enormous"),
                      ordered = TRUE))


dogs <- c("t", "l", "n", "e", "t", "l", "l", "m")
(dog_size_f <- factor(x = dogs,
                      levels = c("m", "t", "n", "l", "e"),
                      labels = c("microscopic", "tiny", "normal", "large", "enormous"),
                      ordered = TRUE))

```

#reading character data from file


```{r}
species <- read.table(file = "~/git_projects/datasets/britain_species/britain_species.txt", header = T, sep = ";", stringsAsFactors = T, as.is = c(3, 5, 6))
str(species)
```
```{r}
weight_classes <- c(3, 2, 3, 1, 4, 2, 3, 2, 4, 1, 1)
(dog_size_f <- factor(x = weight_classes,
                      levels = c(1:5),
                      labels = c("microscopic", "tiny", "normal", "large", "enormous"),
                      ordered = TRUE))
```


## other conversions


```{r}
numbers <- c("1", "3", "-2", "4.99999", "0", "a")
as.integer(numbers)
as.numeric(numbers)
as.logical(numbers)
as.logical(as.numeric(numbers))
x <- c(2, pi, 4)
x + 1
as.character(x) + 1
```



# Selecting from dataframes

```{r}
head(iris, n = 10)
sepal_length <- iris$Sepal.Length
sepal_length[sepal_length < 4.5]

iris[iris$Sepal.Length < 4.5, c("Sepal.Length", "Species")]
iris[c(rep(F, 9), T), c(1:2, 5)]

iris[iris$Petal.Length > 2, ]
iris[iris$Petal.Length == max(iris$Petal.Length), ]
with(iris, {
    sel <- iris[Petal.Length > 5, ]
    table(sel$Species)
       })
subset(x = iris,
       subset = Petal.Length > 5 & Petal.Width < 1.6,
       select = c(3:5))
```

# quantiles

```{r}
quantile(iris$Sepal.Length, probs = c(0, 0.5, 1))

cutoff <- quantile(iris$Sepal.Length, probs = 0.9)
iris$Sepal.Length[iris$Sepal.Length > cutoff]
iris[iris$Sepal.Length > cutoff, ]
iris$Sepal.Length > cutoff
```

```{r}

cutoff <- quantile(body_measurements$age, probs = 0.9)
cutoff


```

## plot multiple series

```{r}
# png(filename = "~/Desktop/myplot.png", 
    # width = 100, 
    # height = 60, 
    # units = "mm",
    # res = 64)
plot(x = iris$Sepal.Length, 
     y = iris$Sepal.Width,
     pch = 20,
     col = "red",
     xlab = "Sepal length (cm)",
     ylab = "Width (cm)",
     ylim = c(0, 5))
points(x = iris$Sepal.Length,
       y = iris$Petal.Width,
       col = "blue",
       pch = 20)
legend("bottomright", 
       legend = c("Sepal", "Petal"), 
       col = c("red", "blue"), 
       pch = 20)
# dev.off()
```

## module and integer division

```{r}
pound_in_grams <- 4536
ounce_in_grams <- 28.35
kilograms <- 100
pounds <- (kilograms * 1000) %/% pound_in_grams
remainder_in_grams <- (kilograms * 1000) %% pound_in_grams
ounces <- remainder_in_grams %/% ounce_in_grams
remainder_in_grams <- remainder_in_grams %% ounce_in_grams

paste0(kilograms, 
       " kg is the same as ",
       pounds, 
       " pounds, ",
       ounces,
       " ounces, and ",
       round(remainder_in_grams, 2),
       " grams")
```



```{r}
age <- 43
if (age > 18) {
  print("you may drive") 
} else {
  print("you sit in the back")
}

```


## iflse functin

```{r}
airquality$label <- ifelse()
  
  
head(airquality)
```

